Supplementary Materials from: \textit{Innovation and elaboration on the avian tree of life}; Science Advances, in review

### 2023-08-21 Manuscript/main.tex
Update main.tex

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Jen A. Bright$^{2}$,
Christopher R. Cooney${^1}$,
Emma C. Hughes${^1}$,
Zo\"{e} K. Varley${^3,4}$,
Natalie Cooper$^{3,+}$,
Andrew P. Beckerman$^{1,+}$,
and Gavin H. Thomas$^{1,4,+}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

@@

The juxtaposition of microevolutionary predictions to macro- and mega-evolutionary observations is striking and raises the question of how phenotypic divergence accumulates across evolutionary scales.
In particular, how can megaevolutionary jumps emerge from gradual micro- or macro-evolutionary processes of phenotypic evolution implied by microevolutionary predictions? \cite{hansen2016,jablonski2022evolvability}
The original descriptions of microevolutionary lines of least resistance explicitly refer to genetic constraints.
However, at the macroevolutionary scale the major axis of phenotypic variation is an emergent property of broader set of genetic and developmental constraints interacting with selection on a moving adaptive landscape % GT: Masahito suggested a whole paragraph on adaptive landscapes and Simpon-Lande-Arnold ideas. I'm not sure that will help but hopefully the expanded intro overall works better. Possibly just needs another sentence or two here but it would break the flow. more citations of relevant papers here (the first 7 or so papers from Gustavo's paper would be relevant.
\cite{BurinWhales,jones2004evolution}. % GT: More references needed? I'm not sure where the idea originates.
At this scale the major axis of phenotypic variation among species \cite{marroig2005size, fasanelli2022allometry} can be estimated from a matrix of divergences in species mean phenotypic traits (e.g. the \textbf{D} matrix of \cite{BlowsHiggie2003, mcglothlin2018adaptive}.
These matrices do not take into account phylogenetic relationships among species.
In contrast, the \textbf{R} matrix of \cite{Houle2017} describes the rate of among-species divergence in species traits and explicitly incorporates phylogeny (also see the \textbf{B} matrix of \cite{Machado2020}) %@L_Rmatrix.
We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}. %TG: TODO: change throughout (that's the VCV matrix now)
The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales. %@L_Rmatrix_full

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/cheat_sheet.pdf}
\caption{\scriptsize{Relationships between key figures in the main text.
A) represents the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
The gray ellipse represents the overall phylogenetic major axis of beak variation in birds.
Species aligned or away from this axis are respectively elaborators or innovators relative to the class Aves.
The red ellipse and axis represents the phylogenetic major axis of beak variation in Tinamiformes only.
Species aligned or away from this axis are respectively elaborators and innovators relative to the Tinamiformes order.
B) In Figure 2, we measured elaboration and innovation at the macroevolutionary scale , i.e. the position of each species in terms of $elaboration_{species}$ and $innovation_{species}$ relative to their order's phylogenetic major axis of beak variation.
A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation(i. e. their position on that axis) whereas their $innovation_{species}$ score is their rejection from it (i.e. their distance away from that axis).
The colours on the tips of the phylogeny correspond to the median $elaboration_{species}$ (orange gradient) and $innovation_{species}$ (blue gradient).
C) Figure 4, we measured elaboration and innovation at the megaevolutionary scale, i.e. $elaboration_{clade}$ and $innovation_{clade}$.
The ellipses are scaled and centered and represent the clade's phylogenetic major axis of beak variation relative to the overall phylogenetic major axis of beak variation with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis of beak variation onto the overall phylogenetic major axis of beak variation and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
The distribution elaboration and innovation scores are taken from the projection of the 4000 pairs of evolutionary rate matrices with each pair being the focal level, e.g. order,and the parent level, e.g. the whole bird phylogeny.}}
\label{Fig:cheatsheet}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/orthogonality_results.pdf}
\caption{\scriptsize{Amount of orthogonality of each clade's phylogenetic major axis of beak variation compared to their parent or parent's parent clade.
The amount of orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background gray and dashed gray lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
Dots represent the median orthogonality of each clade and the lines their 95\% confidence intervals (CI).
No super-order's phylogenetic major axis of beak variation is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal ($>$  $45^\circ$) to the class-wide phylogenetic major axis of beak variation.
These results are to be contrasted with the orientation of each order relative to their super-order where on average (median) only 14/27 are at least half orthogonal to the orientation of their super-orders’s phylogenetic major axis of beak variation.
We also indicate the number of species (n) and the standard deviation (sd) of the orientation of their phylogenetic major axis of beak variation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
Note that noclades are clearly parallel to their parent’s or parent’s parent’s ellipses (i.e.
no 95\% CI includes 0).
For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the class-wide phylogenetic major axis of beak variation relative to their sample size and sd.
The stars represent the posterior probability of the clade's orientation being different from the comparison clade (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
Only 1/6 super-orders, 8/27 orders relative to class-wide phylogeny and 6/27 relative to their super-orders have a pp $> 0.99$.
This is the result of the variation in sample size (n) and standard variation (sd) among clades.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. \ref{fig_orthogonality_passeriformes}.}}
\label{Fig:orthogonality}
\end{figure}
\bigskip

@@

\section*{Reproducibility and repeatability}
The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.

>>>>

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Jen A. Bright$^{2}$,
Christopher R. Cooney${^1}$,
Emma C. Hughes${^1}$,
Zo\"{e} K. Varley${^3,4}$,
Natalie Cooper$^{3,+}$,
Andrew P. Beckerman$^{1,+}$,
and Gavin H. Thomas$^{1,4,+}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\
$^{+}$Contributed equally.\\}

@@

The juxtaposition of microevolutionary predictions to macro- and mega-evolutionary observations is striking and raises the question of how phenotypic divergence accumulates across evolutionary scales.
In particular, how can megaevolutionary jumps emerge from gradual micro- or macro-evolutionary processes of phenotypic evolution implied by microevolutionary predictions? \cite{hansen2016,jablonski2022evolvability}
The original descriptions of microevolutionary lines of least resistance explicitly refer to genetic constraints.
However, at the macroevolutionary scale the major axis of phenotypic variation is an emergent property of broader set of genetic and developmental constraints interacting with selection on a moving adaptive landscape \cite{BurinWhales,jones2004evolution}.
At this scale the major axis of phenotypic variation among species \cite{marroig2005size, fasanelli2022allometry} can be estimated from a matrix of divergences in species mean phenotypic traits (e.g. the \textbf{D} matrix of \cite{BlowsHiggie2003, mcglothlin2018adaptive}.
These matrices do not take into account phylogenetic relationships among species.
In contrast, the \textbf{R} matrix of \cite{Houle2017} describes the rate of among-species divergence in species traits and explicitly incorporates phylogeny (also see the \textbf{B} matrix of \cite{Machado2020}) %@L_Rmatrix.
We hereafter refer to this \textbf{R} matrix as the \textit{evolutionary rate matrix}.
The major axis of the evolutionary rate matrix is a phylogenetic line of least resistance applicable at macro- and mega-evolutionary scales that is analogous to the genetic line of least resistance at microevolutionary scales. %@L_Rmatrix_full

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/cheat_sheet.pdf}
\caption{\scriptsize{Relationships between key figures in the main text.
A) the two first principal components (PCs) of the eight dimensional trait space with all bird beak shapes represented as gray circles and the Tinamiformes as red circles.
The gray ellipse represents the overall phylogenetic major axis of variation.
Species aligned or away from this axis are respectively elaborators or innovators relative to the class Aves.
The red ellipse and axis represents the phylogenetic major axis of variation in Tinamiformes only.
Species aligned or away from this axis are elaborators/innovators relative to the Tinamiformes order.
B) Figure 2 shows elaboration and innovation at the macroevolutionary scale: $elaboration_{species}$ and $innovation_{species}$.
A species $elaboration_{species}$ score is the species projection onto their order's phylogenetic major axis of beak variation (i.e. their position on that axis) whereas their $innovation_{species}$ score is their rejection from it (i.e. their distance away from that axis).
The colours on the tips of the phylogeny correspond to the median $elaboration_{species}$ (orange gradient) and $innovation_{species}$ (blue gradient).
C) Figure 4 shows elaboration and innovation at the megaevolutionary scale: $elaboration_{clade}$ and $innovation_{clade}$.
The ellipses are scaled and centered and represent the clade's phylogenetic major axis relative to the overall phylogenetic major axis with the relative length of the ellipse on each dimension represented on the barplots as variance on each dimension.
A clade's $elaboration_{clade}$ and $innovation_{clade}$ scores are measured by projecting the clade's phylogenetic major axis onto the overall phylogenetic major axis and measuring the projection (elaboration) and rejection (innovation) from this axis as described above.
The distribution elaboration and innovation scores are taken from the projection of the 4000 pairs of evolutionary rate matrices with each pair being the focal level, e.g. order,and the parent level, e.g. the whole bird phylogeny.}}
\label{Fig:cheatsheet}
\end{figure}
\bigskip

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/orthogonality_results.pdf}
\caption{\scriptsize{Amount of orthogonality of each clade's phylogenetic major axis of beak variation compared to their parent or parent's parent clade.
The amount of orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background gray and dashed gray lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
Dots represent the median orthogonality of each clade and the lines their 95\% confidence intervals (CI).
No super-order's phylogenetic major axis is parallel to the class-wide phylogenetic major axis of beak variation (purple lines and circles) and 22/27 orders are on average (median) at least half orthogonal ($>$  $45^\circ$) to the class-wide phylogenetic major axis.
The orientation of each order relative to their super-order where on average (median) at least half orthogonal to the orientation of their super-orders’s phylogenetic major axis of beak variation for 14/27 of the orders.
We also indicate the number of species (n) and the standard deviation (sd; expressed in degrees) of the orientation of their phylogenetic major axis over the 4000 variance-covariance posteriors).
Note that no clades are clearly parallel to their parent’s or parent’s parent’s ellipses (i.e.
no 95\% CI includes 0).
For each clade we measured the posterior probability of each clade's orientation being different from their parent's clade or the class-wide phylogenetic major axis of beak variation relative to their sample size and sd (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
Only 1/6 super-orders, 8/27 orders relative to class-wide phylogeny and 6/27 relative to their super-orders have a pp $> 0.99$.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. \ref{fig_orthogonality_passeriformes}.}}
\label{Fig:orthogonality}
\end{figure}
\bigskip

@@

\section*{Data and Materials Availability}
The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{github} \cite{githubrepo}.
